## Coursera.org 
# Reproducible Research Course   
## Peer Assessment 1



The data for this assignment is available [here] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and includes a total of 17,568 observations.


## 1. Loading and preprocessing the data

After downloading and unzipping dataset file, we can load the data
```{r loading, echo=TRUE}
actData <- read.csv("activity.csv")
```
and prepare 1) a data set **histData** including the total numbers of steps taken each day
```{r histData, echo=TRUE}
histData <- aggregate(steps ~ date, data=actData, FUN=sum)
head(histData)
```
Also we need to create 2) a data set **intervalData** including the means of steps taken per interval
```{r intervalData, echo=TRUE}
intervalData <- aggregate(steps ~ interval, data=actData, FUN=mean)
head(intervalData)
```
These data sets will be useful for us to answer the questions of this assignment. 
Pay attention: the *aggregate()* function has *na.omit* option as default. So we must not do anything with NA-values at this stage.

## 2. What is mean total number of steps taken per day?

Let's plot a histogram of the total number of steps taken each day:
```{r hist_01, echo=TRUE, fig.width=12}
hist(histData$steps, breaks=30, col="grey", border="black", 
     main="Total Number of Steps per Day \nHistogram", xlab="Number of Steps")
abline(v=mean(histData$steps), col="red")
```

We can calculate the mean and median total number of steps taken per day:
```{r means, echo=TRUE}
meanVal <- mean(histData$steps)
medianVal <- median(histData$steps)
```

So **the Mean** = `r signif(meanVal, 5)` , **the Median** = `r signif(medianVal, 5)`  


## 3. What is the average daily activity pattern?

At first we must make some changes in prepared **intervalData**.
What is 5-min interval's format?
```{r interval_format, echo=TRUE}
head(intervalData$interval, 24)
```
We can see the interval values indeed have the format: **hours * 100 + minutes**. It's very comprehensive, but the values are not continuous. So to avoid any distortions in our plot and provide readable time format we need modify interval data.  
1) Transform interval field to get **readable** time format. 
```{r str_intervals, echo=TRUE}
intervalData$interval <- paste(sprintf("%02d", intervalData$interval %/% 100), 
                               sprintf("%02d", intervalData$interval %% 100), sep = ":")
```
2) Add a field of **continuous** interval values:
```{r num_intervals, echo=TRUE}
intervalData <- cbind(seq(1, nrow(intervalData)), intervalData)
names(intervalData)[1] <- "interval.num"
```
Now we can look at the **intervalData**:
```{r interval_data, echo=TRUE}
head(intervalData)
```
**intrval.num** is to be used to make continuous plot and **interval** to mark x-axis by readable time interval values.    
The time series plot of the 5-minute intervals vs the average number of steps is very demonstrative.
```{r time_series, echo=TRUE, fig.width=12}
plot(intervalData$steps  ~ intervalData$interval.num, type="l", main="Average Daily Activity Pattern", 
     xlab="Time Intervals", ylab="Steps", xaxt = "n") ## xaxt = "n" to suppress original x-axis
ticks = seq(1, nrow(intervalData), 24) ## two hours intervals for new axis ticks
axis(side = 1, at = ticks, labels = intervalData$interval[ticks]) ## new x-axis
stepMaxIntrvl <- intervalData$interval.num[which.max(intervalData$steps)]
abline(v=stepMaxIntrvl, col="red")
text(stepMaxIntrvl+10, y = 0, 
     labels=intervalData$interval[intervalData$interval.num == stepMaxIntrvl], col="red")
```

The 5-minute Interval starting at **`r intervalData$interval[intervalData$interval.num == stepMaxIntrvl]`**  contains the maximum number of steps = **`r signif(intervalData$steps[stepMaxIntrvl], 5)`**

## 4. Imputing missing values

Primarily let's test the original data set on NA rows:
```{r, echo=TRUE, results='hide'}
naNum <- sum(is.na(actData))
```
Totale number NA rows = **`r naNum`**

Now let's look at summary of original data set:
```{r nasearch, echo=TRUE}
summary(actData)
```

We can see it's the steps column only contains NAs.  
So let's fill the NA cells with mean values for corresponding 5-minute intervals.  

Primarily we must 1) create set of NA raws ID (*naCells*):  

```{r fullfill, echo=TRUE}
naCells <- which(is.na(actData$steps))
fullActData <- actData
```
2) Transform interval field of fullActData (it will set up the correspondence with intervalData):
```{r set_str_intervals, echo=TRUE}
fullActData$interval <- paste(sprintf("%02d", fullActData$interval %/% 100), 
                               sprintf("%02d", fullActData$interval %% 100), sep = ":")
```

Now we can use loop to substitute NA values with mean value from *intervalData* of the same intervals:
```{r loop, echo=TRUE}
for(i in naCells){
  fullActData$steps[i] <- intervalData$steps[intervalData$interval == fullActData$interval[i]]
}
```
OR we can just repeat values of intervalData 61 times and so again substitute NA values with corresponding mean values:
```{r repeat, echo=TRUE}
avrVals <- rep(intervalData$steps, times=61)
fullActData <- cbind(actData, avrVals)
fullActData$steps[naCells] <- fullActData$avrVals[naCells]
```
The same result. At any cases we have new data set **fullActData** containing means interval values instead of NA.
Let's plot the histogram based on this data set:
```{r filledDataHist, fig.width=12}
histData <- aggregate(steps ~ date, data=fullActData, FUN=sum)
hist(histData$steps, breaks=30, col="grey", border="black", 
     main="Total Number of Steps per Day \nHistogram", xlab="Number of Steps")
abline(v=mean(histData$steps), col="red")
meanValNew <- mean(histData$steps)
medianValNew <- median(histData$steps)
```

And we can see that imputing missing values make the median values of steps number more frequent.  
**The Mean** = `r signif(meanValNew, 6)` , **The Median** = `r signif(medianValNew, 6)`

## 5. Are there differences in activity patterns between weekdays and weekends?

Now I need to set up standard locale for time formats
because the week days names in my locale have another names
```{r localon, echo=TRUE}
sysLoc <- Sys.setlocale(category = "LC_TIME", locale = "C")
```

Define weekday and weekend dates:

```{r weekdays, echo=TRUE}
wDays <- weekdays(as.Date(fullActData$date))
wDays <- gsub("Saturday|Sunday", replacement="weekend", wDays)
wDays <- gsub("Monday|Tuesday|Wednesday|Thursday|Friday", replacement="weekday", wDays)
```

Return to native locale for time formats

```{r localoff, echo=TRUE}
sysLoc <- Sys.setlocale(category = "LC_TIME", locale = "")
```

Set the data for weekends and weekdays:

```{r weekdata, echo=TRUE}
fullActData <- cbind(fullActData, wDays)
weekendData <- fullActData[fullActData$wDays == "weekend",]
weekdayData <- fullActData[fullActData$wDays == "weekday",]
weekendData <- aggregate(steps ~ interval, data=weekendData, FUN=mean)
weekdayData <- aggregate(steps ~ interval, data=weekdayData, FUN=mean)
```


Now again 1) transform interval fiels - *just copy intervalData$interval values*:
```{r str_intervals_again, echo=TRUE}
weekendData$interval <- intervalData$interval
weekdayData$interval <- intervalData$interval
```

and 2) add a field of continuous values for intervals - *just add a copy of intervalData$interval.num values*:
```{r num_intervals_again, echo=TRUE}
weekendData <- cbind(intervalData$interval.num, weekendData)
weekdayData <- cbind(intervalData$interval.num, weekdayData)
names(weekendData)[1] <- "interval.num"
names(weekdayData)[1] <- "interval.num"
```
Plot the data:
```{r weekplot, echo=TRUE, fig.height=12, fig.width=12}
op <- par(mfrow=c(2,1))
ticks = seq(1, nrow(weekdayData), 24) ## two hours intervals for new axis ticks
plot(steps  ~ interval.num, data = weekendData, type="l", main="Average Daily Activity Pattern\n Weekends", 
     xlab="Time intervals", ylab="Steps", xaxt = "n") ## xaxt = "n" to suppress original x-axis
axis(side = 1, at = ticks, labels = weekendData$interval[ticks]) ## new x-axis
plot(steps ~ interval.num, data = weekdayData, type="l", main="Weekdays", 
     xlab="Time Intervals", ylab="Steps", xaxt = "n") ## xaxt = "n" to suppress original x-axis
axis(side = 1, at = ticks, labels = weekdayData$interval[ticks]) ## new x-axis
par(op)
```

So we can see  
1. Weekdays activity max value is more then Weekend one (**`r signif(max(weekdayData$steps), 5)`** vs **`r signif(max(weekendData$steps), 5)`**)  
2. Nevertheless the mean value of Weekend activity is more then mean activity of Weekdays (**`r signif(mean(weekendData$steps), 4)`** vs **`r signif(mean(weekdayData$steps), 4)`**)  
3. The most active intervals are approximately the same at weekdays or weekends. 
